# Two brains or one? {#twoBrains}

We all have a "left brain" and a "right brain", which have different learning styles, and they should be trained both separately and in a way that gets them communicating with each other. Or so many would have you believe. The anatomical division of the hemispheres has been hyped into a supposed plethora of implications for behavior and learning. A veritable industry of education consultants and "brain training" providers have peddled these ideas, selling programs they refer to as, for example, "left brain versus right brain" training [@kroezeBrainGymPseudoscientific2016].

The established science of the matter is that while the much of sensory and perceptual processing is very specific to the two halves of the cortex, more cognitive functions such as declarative memory benefit from a very tight integration. Our conscious experience, too, is quite unified. We experience no discontinuity when the movement of our eyes, or of an object, cause an object to shift from one hemifield, where it is processed predominantly by one hemisphere, to the other hemifield. Communication between the two hemispheres happens rapidly and continuously, and there is no good evidence that exercises designed to insure both hemispheres process stimuli have any benefit for learning. 

In "split-brain" patients, some connections between the hemispheres have been lost (usually not all). Despite the loss of the connections between the cortices of the two hemispheres, such patients can still perform visual search in both hemifields, indicating that both hemispheres have the capacity to do that task. Moreover, for tasks such as searching for a target among many distractor objects, spreading the load by distributing the distractors across the two hemifields can yields a large benefit, suggesting that both hemispheres can search independently and simultaneously [@luckIndependentAttentionalScanning1994a]. But for normal individuals, no such advantage is seen, suggesting that the processes that evaluate each stimulus for whether it is the target are integrated across the hemispheres into a single attentional focus [@luckIndependentHemisphericAttentional1989].

While normal behavior typically indicates a tight integration between the the hemispheres, they do specialize in different functions. In most people, the left hemisphere specializes in language functions such as reading, while the right hemisphere is better at recognizing faces. A behavioral consequence of this is that response times for a face recognition task are slightly faster when the stimulus is presented wholly in the left hemifield (to the right hemisphere) than when it is presented wholly in the right hemifield, whereas the opposite is found for word reading [@rizzolattiOppositeSuperioritiesRight1971]. However, with extended time to process a stimulus, such behavioral asymmetries can disappear as the information from one hemisphere is communicated to the other.

In the performance of most perceptual and attentional tasks, then, in a humans with an intact brain there is little sign that that brain is divided into two halves. Multiple object tracking is, however, a striking exception to this. This is important because it tells us that the limited resource that determines how many objects one can keep track of resides largely with a process that operates independently in the two hemispheres.

## The extraordinary hemifield independence of object tracking

```{r, include=F}
# 
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
#alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
#chance<- 0.5

#Correct for guessing
#What is the true rate, t, at which people know the answer not from guessing, where guessing rate = c
# accuracy = t + (1-t)c
# t = accuracy - (1-t)c
# t = accuracy - c + tc
# t - tc = accuracy - c
# (1-c)t = accuracy - c
# t = (accuracy - c)/(1-c)

#Try deriving it again to make sure I'm right:
# a = t*1 + (1-t)*c
# a = t + c -tc
# t - tc = a - c
# t(1-c) = a - c
# t = (a - c)/(1 - c)

correctForChance <- function(x,chance) {
  #(x-chance)/chance
  (x-chance)/(1-chance)
}
#correctForChance(.89,chance)
library(dplyr)
#1:10 %>% purrr::map(rnorm, n=10)

#accCorrected <- alvarezCavanagh2005E1 %>% purrr::map(correctForChance,chance) %>% 
#  purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector
#accCorrected

calculateIndependence <- function(accuracies,chance) {
  corrected <- accuracies %>% purrr::map(correctForChance,chance) %>%
        purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector

  costOfSecondTargetUnilateral <- corrected[1] - corrected[2]
  # 1 - cost of second target bilateral / costOfSecondTargetUnilateral
  hemifieldIndependence <- 1 - (corrected[3]-corrected[4]) / costOfSecondTargetUnilateral
  return (hemifieldIndependence)
}

```

Evidence for the hemispheric independence of the capacity limit on multiple object tracking was first reported by Alvarez & Cavanagh in 2005. The objects they used resembled spinning pinwheels, an individual bar of which could be designated as a target. Participants' tracking performance was assessed for two targets (one bar of each of two different pinwheels) and for one target [@alvarezIndependentResourcesAttentional2005].  In the two-target condition, when the second target was positioned in the same hemifield as the first target, accuracy was much worse (89% vs. 63%). Remarkably, however, when the second target belonged to a pinwheel in the opposite hemifield, there was very little performance decrement - accuracy was 93% in the one-target condition, and 90% correct in the two-target condition, not a statistically significant difference. This suggests that the processes that limit successful tracking in this task are largely specific to each hemifield. 

Of course, it was already known that the early perceptual processing of targets was independent in the two hemispheres. As was mentioned above, sensory processing and quite a lot of perceptual processing occurs fairly independently in the two hemispheres as the signals from our retinas penetrate into the cortex. What is surprising here is that an ability limited by processing with very limited capacity would show such hemisphere specificity. Such capacities were traditionally thought to be among the processes that are tightly integrated across the two hemispheres, forming a single resource "pool", not two independent pools. We will get back to this point after a quick review of the degree of independence found in tracking.

## Quantitative estimates of independence

```{r, include=F}
#E1: Unilateral 89% 1 target -> 63% 2 target.  Bilateral: 93% 1 target -> 90% 2 target. (26-3)/26= 88%. E3: Unilateral 2targets:93%, 4targets:67%. Bilateral 2targets:90%, 4targets:87%. (26-3)/26=88%
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
alvarezCavanagh2005E3 <- c(.93,.67,.90,.88)
chance<- 0.5

alvarezCavanaghE1Independence <- calculateIndependence(alvarezCavanagh2005E1,0.5)
alvarezCavanaghE1Independence

alvarezCavanaghE3Independence <- calculateIndependence(alvarezCavanagh2005E3,0.5)
alvarezCavanaghE3Independence
```

The degree of hemispheric independence for a task can be estimated. If adding a second stimulus to a hemifield reduces performance by, say, 20 percentage points, but adding that stimulus to the other hemifield reduces performance by only 5 percentage points, one can summarize this as (20-5) / 20 = 75% hemifield independence. A proper calculation of independence is a bit more complicated, however, in that one needs to correct for the accuracy one can achieve by guessing. When I [did the slightly more complication](https://github.com/alexholcombe/tracking-review/blob/main/twoBrainsOrOne.Rmd), the estimated level of independence is very high: `r round(100*alvarezCavanaghE1Independence)`% independence in one of the @alvarezIndependentResourcesAttentional2005 experiments, and `r round(100*alvarezCavanaghE3Independence)`% in the other. 

@alvarezIndependentResourcesAttentional2005 themselves did not report these figures. They calculated expected performance if the hemifields are in fact completely independent, and reported that performance was not statistically significantly worse than that figure. They then suggested that there was complete independence, but this is based on the common fallacy of concluding a null hypothesis is true when the evidence does not reject it at a p<.05 level [@aczelQuantifyingSupportNull2018]. That is, for the null hypothesis they started with their conclusion of complete independence, and then affirmed this conclusion on the basis of not finding much evidence against it. Nevertheless, their data certainly do suggest strong hemispheric independence.

Subsequent work by @chenResourceDemandsObject2013 <!--Didn't compare one to two targets in hemifield but did compare two targets in a hemifield to four targets across two hemifields--> supported the conclusion of a high degree of independence, as did  @holcombeExhaustingAttentionalTracking2012, who used speed thresholds which resulted in a conclusion of about 75% independence. <!--Calculated two-target prediction, and was about 75% independent, but plenty of noise non-significant speed threshold decreases of c(1.67,1.44,1.7,1.63) 1.98rps->1.85 -->.  @shimNumberAttentionalFoci2010<!--Figure 1--> also found a substantial bilateral advantage. 

Interestingly, the finding of statistically significant hemispheric independence has not always replicated [@shimSpatialSeparationTargets2008] <!--performance in all the conditions involved in their comparison were greater than 85% correct, including many low speeds where performance was near ceiling -->  <!--For most of these tasks, but not @shimNumberAttentionalFoci2010, more targets was confounded with having to make more responses and remember said responses at the end of the trial, so some of the residual cost could be due to that being unitary rather than hemifield-specific.-->, but the successful replications, and the substantial statistical power associated with them (evident in their error bars) strongly suggest that at least in some circumstances, tracking does occur mostly independently in the two hemispheres. @shimSpatialSeparationTargets2008 suggested that the reason they did not find evidence for hemifield independence is that they used two targets, whereas the original @alvarezIndependentResourcesAttentional2005 report of hemifield independence used four targets. However, this is unlikely to be the reason for the discrepancy because @holcombeExhaustingAttentionalTracking2012 also used two targets and did find independence. The @shimSpatialSeparationTargets2008 data may have been affected by the high performance level, which is so high that it may have created a ceiling effect. Another possible factor is that the papers findings independence used circular trajectories. However, using trajectories that were not circular @hudsonHemifieldEffectsMultiple2012 found partial independence, although they used a rather different task (multiple identity tracking, which I discuss in a different section).

@carlsonQuadranticDeficitReveals2007 found evidence not only for hemifield independence but also quadrant-level independence, which they attributed to the partial separation of the retinotopic quadrant representations in areas V2 and V3. @shimSpatialSeparationTargets2008 did not replicate this finding, but they had less evidence.

In summary, the evidence for hemifield independence implicates processes that sit at hemisphere-specific processing, or possibly even processing specific to the visual areas that have quadrant-specific retinotopic areas.

## gg

This naturally raises the question of what sort of processes those are. Many authors have chalked this up to spatial interference. They have suggested that inhibitory connections are responsible: when an object is tracked, receptive fields that are nearby on cortical maps are suppressed [@carlsonQuadranticDeficitReveals2007; @franconeriFlexibleCognitiveResources2013]. However, @holcombeObjectTrackingAbsence2014 found evidence against spatial interference extending any further beyond the crowding range, rather than it extending to the limits of a hemifield or quadrant.

An alternative is that something more like the classic concept of an attentional resource can explain hemifield independence. Parietal areas are heavily implicated in tracking, with activation for 

For instance, a population of neurons in parietal areas, which

resource that . a population of neurons in e.g.  parietal areas

with tracked objects

, which naturally are less likely to cross hemifield boundaries because of the greater distance 

that are less likely to cross hemifields or 

attributed this to spatial interference being 

nice side effect of this discovery

Of course, we already knew that visual cortex

This is noteworthy because

A great thing about this discovery is that it shows decisively that MOT processes are differentiable from more cognitive processes.


provides a new tool

The calculation rests on assumptions that are not secure. That when given the task of tracking two targets, if you can only track one, you still do just as well on the first, while guessing on the second

We easily attend to objects in any part of the visual field, left or right. 

The hemispheric division of cortex also means that at least in early visual cortex, stimuli presented in one visual field are processed independently of stimuli presented simultaneously in the other visual field. This has a striking effect on behavioral performance of one task in particular.


<!-- Shim, Alvarez, & Jiang (2008) found no advantage for targets in different hemifields. Why? Not sure, but the objects weren’t spaced far enough part to not crowd each other across the hemifield boundaries. In contrast, Shim et al. and Bello et al. suggest it was because hemifield limit doesn’t apply when just two targets. Holcombe & Chen found it did, but Bello suggests the reason is that we used circular trajectories that are somehow different, they say you can extrapolate more with them, whereas it may be that you can extrapolate less. -->


The target-target distance effect in Shim et al. is huge! How do I explain that? One possible explanation is Carlson’s quadrantic deficit.
Most cognitive tasks are unlikely to be independently mediated by the two hemispheres, with one set of processes in each hemisphere.
Even feature attention is global @liuConstantSpreadFeaturebased2011

have such a high level of hemispheric independence.




@chenResourceDemandsObject2013 This unitary resource doesn't do that much in tracking!

corroborated this conclusion that the two hemispheres

<!-- RT study, always used 2 targets.  Davis, R., & Schmit, V. (1971). Timing the transfer of information between hemispheres in man. Acta Psychologica, 35(5), 335–346. https://doi.org/10.1016/0001-6918(71)90008-4 -->


@chenResourceDemandsObject2013 @croweGoaldirectedUnequalAttention2019


For most tasks, however, there seems to be little advantage to 

if some stimuli are presented in one visual field

There is another subtle effect of the hemispheric division of cortex. In some tasks involving two simultaneously-presented stimuli, higher performance is found when each stimulus is presented in a different hemifield rather then presenting them both in the same hemifield. For example, the reporting of two briefly-presented digits is more accurate when the digits are presented in different hemifields than in the same hemifield [@dimondUseTwoCerebral1971].  




However, many studies did not include a single-stimulus condition, so when performance is higher in the split condition, we don't know how close that is to the one-target level of performance [@awhEvidenceSplitAttentional2000; @serenoDiscriminationHemifieldsNew1991; @dimondUseTwoCerebral1971], or they used response time as a measure, which is especially difficult to interpret.

A complication in calculating capacity is that it assumes that if you can only track one, then if you add a second target you will succeed just as often in tracking one of them. However, it may be that if you split your tracking resource, allocating half to one target and half to another, then you will fail at both. Based both on experience in the tasks, and the evidence provided by @chenResourceDemandsObject2013, in at least some circumstances this appears to be the case. In the jargon, this is to say that the resource function that relates the proportion of attentional resource to accuracy [@normanDatalimitedResourcelimitedProcesses1975] falls below a straight line.
@holcombeSplittingAttentionReduces2013

The situation is diffrent for some tasks that use more briefly-presented stimuli.








, such that presenting a word, for example, yields slightly greater accuracy 

In normal individuals, there are some tasks for which there is evidence of some independent functioning, in the form of a behavioural advantage for presenting the relevant stimuli split across the hemifields rather than presenting them to one hemifield. An early demonstration of this was by 

normal individuals act much more as if their hemispheres are working independently.

<!-- A' is no good https://sites.google.com/a/mtu.edu/whynotaprime/ -->
@delvenneCapacityVisualShortterm2005 used A', which sucks, observed approximately a 40% independence for spatial VSTM <!--   -->.

In visual short-term memory tasks, people perform a good deal better when the stimuli are spread across both the left and right hemifields rather than all presented to the left hemifield or to the right hemifield [@delvenneVisualShorttermMemory2012]. The reason, it is believed, is that there is some sort of resource that is specific to each hemifield. It is better, then, to spread one's load, rather than burdening one hemisphere more than the other. If the processes that limit memory performance were entirely specific to the hemispheres, then once people's limit for remembering objects presented to one hemifield were reached, one could add more objects to the other hemifield without any cost for the proportion of objects remembered. Instead, though, the hemifield advantage is only partial. There are a few tasks, however, for which the hemifield advantage seems to be complete, and multiple object tracking is one.

@delvenneCapacityVisualShortterm2005: "a color and spatial location change detection task, in which the items were displayed either in the two visual fields or in the same hemifield. The data revealed that only memory capacity for spatial locations and not colors increased when the items were separated between the two visual fields. These findings support the view of VSTM as a chain of capacity limited operations where the spatial selection of stimuli, which dominates in both spatial location VSTM and MOT, occupies the first place and shows independence between the two fields." also null effect for color XuNakayama2007

@delvenneBilateralFieldAdvantage2011 showed a bilateral advantage for subitizing. <!-- Subitizng and enumeration. Bilateral advantage shown for enumeration. None for subitizing, although that's hard to interpret because ceiling effect -->

@delvenneVisualShorttermMemory2012 review article: "Those data reveal that a BFA can be found in spatial short-term memory, namely when the task requires holding spatial information. However, no BFA has been found in object short-term memory (i.e., when the task is to hold detailed visual properties of the objects), unless attentional selective processes are dominantly engaged in the task, s"

@umemotoBilateralAdvantageStorage2010 found bilateral advantage even with sequential so must be storage, althoguh used orientation, so could be spatial still
<!-- used mixture model, but  might have assessed capacity quantitatively--> Mixture model suggests only a modest increase, from perhaps capacity of 2 <!--0.5*4, Figure 11--> to capacity of 2.3 <!--.575*4, Figure11-->

@hudsonHemifieldEffectsMultiple2012 multiple identity tracking , found a hemifield advantage: " Contrary to expectations, a bilateral advantage was still observed, though it was not as strong as when observers were not required to remember the identities of the targets. This finding is inconsistent with the only model of multiple identity tracking (Oksama & Hyönä, 2008, Cognitive Psychology, 56, 237-283), so we present an alternative account."

@strongHemifieldspecificControlSpatial2019

RIGHT HEMIFIELD IS BETTER @holcombeObjectTrackingAbsence2014 (Figure A2), Battelli

## Why

Early work had concluded that MOT could not reflect serial position sampling @pylyshynTrackingMultipleIndependent1988; @yantisMultielementVisualTracking1992

It is well-known that a decrease with target load, linear or non-linear, can be explained by parallel models as well as serial models.

It happens for spatial VSTM but not color [@delvenneCapacityVisualShortterm2005]

### Quadrants

Carlson et al. found evidence for quadrants. Has this been replicated? And @strongHemifieldspecificControlSpatial2019 found a cost for crossing the vertical midline but not the horizontal midline. So now we know that the cross-hemisphere cost does not occur for quadrants. 

Hughes & Zimba, 1987
