# Two brains or one? {#twoBrains}

Most of the public know little about the brain, but have heard the phrases "left brain" and "right brain". Understandably, the public assumes that this separation has major consequences for learning. This assumption is exploited by a major industry of pseudoscientific education consultants and brain training programs, with programs they refer to as "left brain versus right brain" training [@kroezeBrainGymPseudoscientific2016] and the like.

The truth is that while early sensory processing is very specific to the two halves of the cortex, more cognitive functions such as explicit learning benefit from a very tight integration. Our conscious experience is one of attending easily to objects in any part of the visual field, left or right. We experience no discontinuity when the movement of our eyes, or of an object, cause an object to shift from one hemifield, where it is processed predominantly by one hemisphere, to the other hemifield. Communication between the two hemispheres happens rapidly and continuously, and there is no good evidence that exercises designed to insure both hemispheres process stimuli have any benefit for learning. 

For those extraordinarily rare individuals who have had some connections between their hemispheres cut ("split brain" patients), the hemispheres act more independently. Indeed, for tasks such as searching for a target among many distractor objects, it is better to distribute the distractors across the two hemifields. Spreading the load in this way yields a large benefit for search, suggesting that both hemispheres can search independently and simultaneously [@luckIndependentAttentionalScanning1994a]. But for normal individuals, no such advantage is seen, suggesting that the processes that evaluate each stimulus for whether it is the target are well integrated across the hemispheres [@luckIndependentHemisphericAttentional1989].

This is not to say that the hemispheres are equivalent. They do specialize in different functions. For example, the left hemisphere in most people seems to specialize in language functions including reading, whereas the right hemisphere is better at recognizing faces. A behavioral consequence is that response times for a face recognition task are slightly faster when the face is presented in the left hemifield (to the right hemisphere) whereas the opposite is found for word reading [@rizzolattiOppositeSuperioritiesRight1971]. However, with extended time to process a stimulus, these behavioral asymmetries, already small, diminish and are hard to detect, as the two hemispheres collaborate on processing a stimulus presented to either hemifield.

The hemispheric division of cortex also means that at least in early visual cortex, stimuli presented in one visual field are processed independently of stimuli presented simultaneously in the other visual field. This has a striking effect on behavioral performance of one task in particular.

## The extraordinary hemifield independence of object tracking

```{r, include=F}
# 
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
#alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
#chance<- 0.5

#Correct for guessing
#What is the true rate, t, at which people know the answer not from guessing, where guessing rate = c
# accuracy = t + (1-t)c
# t = accuracy - (1-t)c
# t = accuracy - c + tc
# t - tc = accuracy - c
# (1-c)t = accuracy - c
# t = (accuracy - c)/(1-c)

#Try deriving it again to make sure I'm right:
# a = t*1 + (1-t)*c
# a = t + c -tc
# t - tc = a - c
# t(1-c) = a - c
# t = (a - c)/(1 - c)

correctForChance <- function(x,chance) {
  #(x-chance)/chance
  (x-chance)/(1-chance)
}
#correctForChance(.89,chance)
library(dplyr)
#1:10 %>% purrr::map(rnorm, n=10)

#accCorrected <- alvarezCavanagh2005E1 %>% purrr::map(correctForChance,chance) %>% 
#  purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector
#accCorrected

calculateIndependence <- function(accuracies,chance) {
  corrected <- accuracies %>% purrr::map(correctForChance,chance) %>%
        purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector

  costOfSecondTargetUnilateral <- corrected[1] - corrected[2]
  # 1 - cost of second target bilateral / costOfSecondTargetUnilateral
  hemifieldIndependence <- 1 - (corrected[3]-corrected[4]) / costOfSecondTargetUnilateral
  return (hemifieldIndependence)
}

```

In 2005, Alvarez & Cavanagh compared performance when participants tracked either one target or two targets  [@alvarezIndependentResourcesAttentional2005]. In a condition where the second target was located in the same hemifield as the first target, accuracy was much worse (89% vs. 63%). Remarkably, however, when the second target was in the opposite hemifield, there was very little performance decrement (93% vs. 90%).  Evidently, the processes that limit successful tracking are largely specific to the two hemifields, suggesting that there are two sets of processes that track multiple objects, one in the left cerebral hemisphere, and one in the right.
  <!--E1: Unilateral 89% 1 target -> 63% 2 target.  Bilateral: 93% 1 target -> 90% 2 target. (26-3)/26= 88%. E3: Unilateral 2targets:93%, 4targets:67%. Bilateral 2targets:90%, 4targets:87%. (26-3)/26=88% -->

```{r, include=F}
# 
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
alvarezCavanagh2005E3 <- c(.93,.67,.90,.88)
chance<- 0.5

alvarezCavanaghE1Independence <- calculateIndependence(alvarezCavanagh2005E1,0.5)
alvarezCavanaghE1Independence

alvarezCavanaghE3Independence <- calculateIndependence(alvarezCavanagh2005E3,0.5)
alvarezCavanaghE3Independence
```

The level of hemispheric independence for a task can be quantified from the advantage provided by presenting the relevant stimuli in distinct hemifields. If adding a second stimulus to a hemifield reduces performance by, say, 20 percentage points, but adding that second stimulus to the other hemifield instead reduces performance by only 5 percentage points, one might say that there is (20-5) / 20 = 75% hemifield independence. That illustrates the principle, but it is not quite this simple, in that one needs to first correct for guessing before interpreting the performance figures.

The result was a truly extraordinary finding: `r round(100*alvarezCavanaghE1Independence)`% independence in one experiment, and `r round(100*alvarezCavanaghE3Independence)`% in a second experiment. 

@alvarezIndependentResourcesAttentional2005 themselves did not report these figures. They did calculate expected performance if the hemifields are in fact completely independent, and found that performance was not statistically significantly worse than that figure. They then suggested that there was complete independence, but this is the common fallacy of concluding a null hypothesis is true when the evidence does not reject it at a p<.05 level. In other words, they assumed (as the null hypothesis) their conclusion of complete independence, and then affirmed the conclusion on the basis of not finding much evidence against it.

The calculation rests on assumptions that are not secure. That when given the task of tracking two targets, if you can only track one, you still do just as well on the first, while guessing on the second

<!--Still, they did find that distributing objects to two hemifields could almost double capacity, as if for this task, the two hemispheres functioned almost as two different brains. -->Subsequent work by @holcombeExhaustingAttentionalTracking2012 <!--Calculated two-target prediction, and was about 75% independent, but plenty of noise non-significant speed threshold decreases of c(1.67,1.44,1.7,1.63) 1.98rps->1.85 --> and @chenResourceDemandsObject2013 found that tracking in the two hemifields was largely independent, but perhaps not completely independent. @shimNumberAttentionalFoci2010<!--Figure 1--> also found a substantial bilateral advantage. This result has not always replicated [@shimSpatialSeparationTargets2008] <!--performance in all the conditions involved in their comparison were great- er than 85% correct, including many low speeds where performance was near ceiling -->  <!--For most of these tasks, but not @shimNumberAttentionalFoci2010, more targets was confounded with having to make more responses and remember said responses at the end of the trial, so some of the residual cost could be due to that being unitary rather than hemifield-specific.-->, but the successful replications, and the substantial statistical power associated with them (evident in their error bars) indicate that hemispheric near-independence of tracking does in fact occur.

Shim, Alvarez, & Jiang (2008) found no advantage for targets in different hemifields. Why? Not sure, but the objects weren’t spaced far enough part to not crowd each other across the hemifield boundaries. In contrast, Shim et al. and Bello et al. suggest it was because hemifield limit doesn’t apply when just two targets. Holcombe & Chen found it did, but Bello suggests the reason is that we used circular trajectories that are somehow different, they say you can extrapolate more with them, whereas it may be that you can extrapolate less.

The target-target distance effect in Shim et al. is huge! How do I explain that? One possible explanation is Carlson’s quadrantic deficit.
Most cognitive tasks are unlikely to be independently mediated by the two hemispheres, with one set of processes in each hemisphere.
Even feature attention is global @liuConstantSpreadFeaturebased2011

have such a high level of hemispheric independence.

likely do not have this independence. 
do not appear to have this independence. For e

A great thing about this discovery is that it shows decisively that MOT processes are differentiable from more cognitive processes.
provides a new tool

A major complication to calculating the amount of hemifield independence in multiple object tracking is that 

While finding hemifield indpendence
@chenResourceDemandsObject2013 This unitary resource doesn't do that much in tracking!

corroborated this conclusion that the two hemispheres

<!-- RT study, always used 2 targets.  Davis, R., & Schmit, V. (1971). Timing the transfer of information between hemispheres in man. Acta Psychologica, 35(5), 335–346. https://doi.org/10.1016/0001-6918(71)90008-4 -->





@chenResourceDemandsObject2013 @croweGoaldirectedUnequalAttention2019


For most tasks, however, there seems to be little advantage to 

if some stimuli are presented in one visual field

There is another subtle effect of the hemispheric division of cortex. In some tasks involving two simultaneously-presented stimuli, higher performance is found when each stimulus is presented in a different hemifield rather then presenting them both in the same hemifield. For example, the reporting of two briefly-presented digits is more accurate when the digits are presented in different hemifields than in the same hemifield [@dimondUseTwoCerebral1971].  




However, many studies did not include a single-stimulus condition, so when performance is higher in the split condition, we don't know how close that is to the one-target level of performance [@awhEvidenceSplitAttentional2000; @serenoDiscriminationHemifieldsNew1991; @dimondUseTwoCerebral1971], or they used response time as a measure, which is especially difficult to interpret.

A complication in calculating capacity is that it assumes that if you can only track one, then if you add a second target you will succeed just as often in tracking one of them. However, it may be that if you split your tracking resource, allocating half to one target and half to another, then you will fail at both. Based both on experience in the tasks, and the evidence provided by @chenResourceDemandsObject2013, in at least some circumstances this appears to be the case. In the jargon, this is to say that the resource function that relates the proportion of attentional resource to accuracy [@normanDatalimitedResourcelimitedProcesses1975] falls below a straight line.
@holcombeSplittingAttentionReduces2013

The situation is diffrent for some tasks that use more briefly-presented stimuli.








, such that presenting a word, for example, yields slightly greater accuracy 

In normal individuals, there are some tasks for which there is evidence of some independent functioning, in the form of a behavioural advantage for presenting the relevant stimuli split across the hemifields rather than presenting them to one hemifield. An early demonstration of this was by 

normal individuals act much more as if their hemispheres are working independently.

<!-- A' is no good https://sites.google.com/a/mtu.edu/whynotaprime/ -->
@delvenneCapacityVisualShortterm2005 used A', which sucks, observed approximately a 40% independence for spatial VSTM <!--   -->.

In visual short-term memory tasks, people perform a good deal better when the stimuli are spread across both the left and right hemifields rather than all presented to the left hemifield or to the right hemifield [@delvenneVisualShorttermMemory2012]. The reason, it is believed, is that there is some sort of resource that is specific to each hemifield. It is better, then, to spread one's load, rather than burdening one hemisphere more than the other. If the processes that limit memory performance were entirely specific to the hemispheres, then once people's limit for remembering objects presented to one hemifield were reached, one could add more objects to the other hemifield without any cost for the proportion of objects remembered. Instead, though, the hemifield advantage is only partial. There are a few tasks, however, for which the hemifield advantage seems to be complete, and multiple object tracking is one.

@delvenneCapacityVisualShortterm2005: "a color and spatial location change detection task, in which the items were displayed either in the two visual fields or in the same hemifield. The data revealed that only memory capacity for spatial locations and not colors increased when the items were separated between the two visual fields. These findings support the view of VSTM as a chain of capacity limited operations where the spatial selection of stimuli, which dominates in both spatial location VSTM and MOT, occupies the first place and shows independence between the two fields." also null effect for color XuNakayama2007

@delvenneBilateralFieldAdvantage2011 showed a bilateral advantage for subitizing. <!-- Subitizng and enumeration. Bilateral advantage shown for enumeration. None for subitizing, although that's hard to interpret because ceiling effect -->

@delvenneVisualShorttermMemory2012 review article: "Those data reveal that a BFA can be found in spatial short-term memory, namely when the task requires holding spatial information. However, no BFA has been found in object short-term memory (i.e., when the task is to hold detailed visual properties of the objects), unless attentional selective processes are dominantly engaged in the task, s"

@umemotoBilateralAdvantageStorage2010 found bilateral advantage even with sequential so must be storage, althoguh used orientation, so could be spatial still
<!-- used mixture model, but  might have assessed capacity quantitatively--> Mixture model suggests only a modest increase, from perhaps capacity of 2 <!--0.5*4, Figure 11--> to capacity of 2.3 <!--.575*4, Figure11-->

@hudsonHemifieldEffectsMultiple2012 multiple identity tracking , found a hemifield advantage: " Contrary to expectations, a bilateral advantage was still observed, though it was not as strong as when observers were not required to remember the identities of the targets. This finding is inconsistent with the only model of multiple identity tracking (Oksama & Hyönä, 2008, Cognitive Psychology, 56, 237-283), so we present an alternative account."

@strongHemifieldspecificControlSpatial2019

RIGHT HEMIFIELD IS BETTER @holcombeObjectTrackingAbsence2014 (Figure A2), Battelli

## Why

It happens for spatial VSTM but not color [@delvenneCapacityVisualShortterm2005]

### Quadrants

Carlson et al. found evidence for quadrants. Has this been replicated? And @strongHemifieldspecificControlSpatial2019 found a cost for crossing the vertical midline but not the horizontal midline. So now we know that the cross-hemisphere cost does not occur for quadrants. 

Hughes & Zimba, 1987
