# Two brains or one? {#twoBrains}

A human brain has two halves, a left and a right, that are anatomically connected, but there are fewer cross-hemisphere connections than there are within-hemisphere connections. 
<!--While an industry of education consultants and "brain training" companies exploits  common left brain, right brain misconceptions,  
We all have a "left brain" and a "right brain", which have different learning styles, and they should be trained both separately and in a way that gets them communicating with each other. Or so many would have you believe. The anatomical division of the hemispheres has been hyped into a supposed plethora of implications for behavior and learning. A veritable industry of education consultants and "brain training" providers have peddled these ideas, selling programs they refer to as, for example, "left brain versus right brain" training [@kroezeBrainGymPseudoscientific2016].--> Much of sensory and perceptual processing runs rather independently in the two halves of the cortex, but more cognitive functions such as declarative memory benefit from a very tight integration. This integration is extensive enough that the comparison of our two hemispheres to our two hands or our two legs is misleading.

Our conscious experience, too, is quite unified. We experience no discontinuity when the movement of our eyes, or of an object, cause an object to shift from one hemifield, where it is processed predominantly by one hemisphere, to the other hemifield. Communication between the two hemispheres happens rapidly and continuously. Despite the claims of salespeople who prey on well-meaning schools and parents, there is no good evidence that exercises designed to insure both hemispheres process stimuli have any benefit for learning. 

In "split-brain" patients, many of the connections between the hemispheres have been lost. Yet such patients can still perform tasks such as visual search in both hemifields, suggesting that both hemispheres have the mechanisms needed to do such tasks. When split-brain patients are asked to search for a target among many distractor objects, spreading the load by distributing the distractors across the two hemifields can yield a large benefit, suggesting that the two hemispheres in these patients carry out their searches independently [@luckIndependentAttentionalScanning1994]. For intact individuals, no such advantage is seen, suggesting that in a normal brain, the processes that evaluate each stimulus for whether it is the target are integrated across the hemispheres into a single attentional focus [@luckIndependentHemisphericAttentional1989].

Although the hemispheres show tight integration during many tasks, each hemisphere does specialize in certain functions. The left hemisphere has greater competence in language functions such as reading, while the right hemisphere is better at recognizing faces. One behavioral consequence is that response times for a face recognition task are slightly faster when the stimulus is presented wholly in the left hemifield (to the right hemisphere) than when it is presented wholly in the right hemifield, whereas the opposite is found for word reading [@rizzolattiOppositeSuperioritiesRight1971]. With extended time to process a stimulus, however, such behavioral asymmetries can disappear, because eventually the information from one hemisphere gets communicated to the other.

From the performance of most perceptual and attentional tasks, then, in typically-developing humans there is little overt sign that the brain is divided into two halves. Multiple object tracking, however, is a major exception to this. The pattern of performance found indicates that the limited resource that determines how many objects one can keep track of resides largely with a process that operates independently in the two hemispheres.

## The extraordinary hemifield independence of object tracking

```{r, include=F}

#Correct for guessing. Based on 1, the chance rate for getting the task of identifying the targets correct.
# And 2, the observed percent correct.
# Use a high-threshold equation that assumes you either know it or you don't.

#What is the true rate, t, at which people know the answer not from guessing, where guessing rate = c
# accuracy = t + (1-t)c
# t = accuracy - (1-t)c
# t = accuracy - c + tc
# t - tc = accuracy - c
# (1-c)t = accuracy - c
# t = (accuracy - c)/(1-c)

#Try deriving it again to make sure I'm right:
# a = t*1 + (1-t)*c
# a = t + c -tc
# t - tc = a - c
# t(1-c) = a - c
# t = (a - c)/(1 - c)

correctForChance <- function(x,chance) {
  #(x-chance)/chance
  (x-chance)/(1-chance)
}
#correctForChance(.89,chance)


# Alvarez and Cavanagh arrayed 2 pinwheels either unilaterally or bilaterally, and in each case designated either one or both as targets
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
# 
#alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
#chance<- 0.5

library(dplyr)
#1:10 %>% purrr::map(rnorm, n=10)

#accCorrected <- alvarezCavanagh2005E1 %>% purrr::map(correctForChance,chance) %>% 
#  purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector
#accCorrected

calculateIndependenceAlvarezCavanagh <- function(accuracies,chance) {
  corrected <- accuracies %>% purrr::map(correctForChance,chance) %>%
        purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector

  costOfSecondTargetUnilateral <- corrected[1] - corrected[2]
  # 1 - cost of second target bilateral / costOfSecondTargetUnilateral
  hemifieldIndependence <- 1 - (corrected[3]-corrected[4]) / costOfSecondTargetUnilateral
  return (hemifieldIndependence)
}

# The HudsonHoweLittle2012 design compares accuracy with 3 targets to with 6 targets
# Instead of having two different baseline (3-targets in their case) conditions like Alvarez & Cavanagh, they have just one, where
# the three targets are confined to a single quadrant and for the 6-target conditions, the 3 additional targets are added to either the other hemifield (bilateral)
# or the same hemifield but other quadrant (unilateral)

calculateIndependenceOnlyOneBaselineConfiguration <- function(accuracies,chance) {
  corrected <- accuracies %>% purrr::map(correctForChance,chance) %>%
        purrr::map_dbl(mean) #have to call map_dbl to simplify output from nested list to simple vector

  costOfAdditionalTargetsUnilateral <- corrected[1] - corrected[2]
  costOfAdditionalTargetsBilateral <- corrected[1] - corrected[3]
  # 1 - cost of second target bilateral / costOfSecondTargetUnilateral
  hemifieldIndependence <- 1 - costOfAdditionalTargetsBilateral / costOfAdditionalTargetsUnilateral
  return (hemifieldIndependence)
}

```

In 2005, George Alvarez & Patrick Cavanagh reported a stunning finding from their MOT task. Using objects that resembled spinning pinwheels, as a target they designated an individual bar of a pinwheel. Performance in a two-target condition (the targets were one bar of each of two different pinwheels) was contrasted with that for a one-target condition [@alvarezIndependentResourcesAttentional2005]. When the second target was positioned in the same hemifield as the first target, accuracy in the two-target condition was much worse (89% vs. 63%). Remarkably, however, when the second target belonged to a pinwheel located in the opposite hemifield, there was very little performance decrement - accuracy was 93% in the one-target condition, and 90% correct in the two-target condition. This suggests that the processes that limit successful tracking in this task are largely specific to each hemifield. 

It was already known that sensory processing and quite a lot of perceptual processing occurs independently in each hemisphere. What is somewhat surprising is that a higher-level, limited-capacity process would be hemisphere-independent. Such capacities were traditionally thought to be among the processes that are tightly integrated across the two hemispheres, forming a single resource "pool", not two independent limits. We will get back to this point, first we'll examine more extensively the evidence for hemispheric independence of object tracking.

## Quantitative estimates of independence

```{r, include=F}
#E1: Unilateral 89% 1 target -> 63% 2 target.  Bilateral: 93% 1 target -> 90% 2 target. (26-3)/26= 88%. E3: Unilateral 2targets:93%, 4targets:67%. Bilateral 2targets:90%, 4targets:87%. (26-3)/26=88%
# Unilateral 1 target, Unilateral 2 target, Bilateral 1 target, Bilateral 2 target
alvarezCavanagh2005E1 <- c(.89,.63,.93,.90)
alvarezCavanagh2005E3 <- c(.93,.67,.90,.88)
chance<- 0.5

alvarezCavanaghE1Independence <- calculateIndependenceAlvarezCavanagh(alvarezCavanagh2005E1,0.5)
alvarezCavanaghE1Independence

alvarezCavanaghE3Independence <- calculateIndependenceAlvarezCavanagh(alvarezCavanagh2005E3,0.5)
```

```{r HudsonHoweLittle, include=F}
# Baseline unilateral 3 targets, Unilateral 6 targets, Bilateral 6 targets
HudHoweLittleE1 <- c(.77,.54,.61)
#Two yes/no questions were asked per trial "Was this color target at this position?" and both had to be answered correctly
chance<- 0.25

HudHoweLittleE1b <- c(.72,.51,.57) #Replication of E1 with eyetracker
#In E1 and E1b the colors were continuously visible.
#E2 the colors were only briefly shown at the beginning of the trial
HudHoweLittleE2 <- c(.75,.52,.58)

#Used same num targets and distractors as Alvarez & Cavanagh: 2 targets and 2 distractors
HudHoweLittleE3 <- c(.77,.47,.58)

#In E4 they used an MOT task
HudHoweLittleMOT <- c(.72,.36,.595)
HudHoweLittleMOTindependence<- calculateIndependenceOnlyOneBaselineConfiguration(HudHoweLittleMOT, chance)

HudHoweLittleE1Independence <- calculateIndependenceOnlyOneBaselineConfiguration(HudHoweLittleE1, chance)
HudHoweLittleE1bIndependence <- calculateIndependenceOnlyOneBaselineConfiguration(HudHoweLittleE1b, chance)
HudHoweLittleE2Independence <- calculateIndependenceOnlyOneBaselineConfiguration(HudHoweLittleE2, chance)
HudHoweLittleE3Independence <- calculateIndependenceOnlyOneBaselineConfiguration(HudHoweLittleE3, chance)
HudHoweLittleMITindependence<- c(HudHoweLittleE1Independence,HudHoweLittleE1bIndependence,HudHoweLittleE2Independence,HudHoweLittleE3Independence)
```

The hemispheric independence of a task can be quantified. Imagine that adding a second stimulus to a hemifield reduces performance by 20 percentage points, but adding that stimulus to the other hemifield reduces performance by only 5 percentage points. One can quantify the hemispheric independence, then, as (20-5) / 20 = 75% hemifield independence. Ideally, however, one would not use raw accuracy but instead would correct for the accuracy one can achieve by guessing. When [applying such a calculation](https://github.com/alexholcombe/tracking-review/blob/main/twoBrainsOrOne.Rmd) to the @alvarezIndependentResourcesAttentional2005 results, the estimated level of independence is very high: `r round(100*alvarezCavanaghE1Independence)`% independence in one of their experiments, and `r round(100*alvarezCavanaghE3Independence)`% in the other. 

@alvarezIndependentResourcesAttentional2005 themselves, like other researchers who have investigated this question, did not do these calculations. @alvarezIndependentResourcesAttentional2005 calculated expected performance if the hemifields are in fact completely independent, and reported that performance was not statistically significantly worse than that figure. They then suggested that there was complete independence, but this is based on the common fallacy of concluding a null hypothesis is true when the evidence does not reject it at a p<.05 level [@aczelQuantifyingSupportNull2018]. That is, for the null hypothesis they inappropriately started with their conclusion (complete independence), and then affirmed this conclusion on the basis of not finding much evidence against it. Nevertheless, their data do suggest hemispheric independence of approximately 90%. In a study with similar methods, @hudsonHemifieldEffectsMultiple2012 found `r round(100*HudHoweLittleMOTindependence)`% independence (they did not calculate a number, so this is [my calculation](https://github.com/alexholcombe/tracking-review/blob/main/twoBrainsOrOne.Rmd)).

Some of the follow-up studies in this area have not included enough conditions to quantify the degree of independence, or confounded distribution of the targets to two hemifields with greater distance among them, such that any benefit might have been due to less spatial crowding interference, a phenomenon discussed in Chapter \@ref(spatialInterference).

@holcombeExhaustingAttentionalTracking2012 and @chenResourceDemandsObject2013 also found evidence for a high degree of independence, using a slightly different approach based on speed thresholds. The findings were compatible with approximately 100% hemifield independence or a bit less. @shimNumberAttentionalFoci2010<!--Figure 1--> and @stormerWithinhemifieldCompetitionEarly2014 also found evidence for a substantial bilateral advantage compared to adding a target in the same hemifield. 

These findings of hemispheric independence have not replicated in all circumstances [e.g., @shimSpatialSeparationTargets2008] <!--performance in all the conditions involved in their comparison were greater than 85% correct, including many low speeds where performance was near ceiling -->  <!--For most of these tasks, but not @shimNumberAttentionalFoci2010, more targets was confounded with having to make more responses and remember said responses at the end of the trial, so some of the residual cost could be due to that being unitary rather than hemifield-specific.-->, but the successful replications strongly suggest that at least in some circumstances, tracking does occur mostly independently in the two hemispheres. I say "mostly independently" rather than suggesting complete independence because each individual study has too much statistical uncertainty to rule out a figure such as 75% independence, even when the point estimates I've reported above suggest a higher degree of independence.

@shimSpatialSeparationTargets2008 suggested that the reason they did not find evidence for hemifield independence is that they used only two targets, whereas according to them the original @alvarezIndependentResourcesAttentional2005 report of hemifield independence used four targets. This is unlikely to be the reason for the discrepancy, however, because in their E1 and E2 @alvarezIndependentResourcesAttentional2005 did find evidence for hemifield independence using just two targets, as did @holcombeExhaustingAttentionalTracking2012 and @stormerWithinhemifieldCompetitionEarly2014. The @shimSpatialSeparationTargets2008 data may have been afflicted by a ceiling effect, as accuracy was over 85% correct in all conditions in their experiment.

<!-- Shim, Alvarez, & Jiang (2008) found no advantage for targets in different hemifields. Why? Not sure, but the objects weren’t spaced far enough part to not crowd each other across the hemifield boundaries. In contrast, Shim et al. and Bello et al. suggest it was because hemifield limit doesn’t apply when just two targets. Holcombe & Chen found it did, but Bello suggests the reason is that we used circular trajectories that are somehow different, they say you can extrapolate more with them, whereas it may be that you can extrapolate less.   The target-target distance effect in Shim et al. is huge! How do I explain that? One possible explanation is Carlson’s quadrantic deficit.-->

A limitation of deriving hemispheric independence from accuracy is that they depend on the assumption that if a person can only track one target, in a condition where the person is also trying to track a second target, the person will succeed just as often in tracking one of the two. My introspective experience, however, indicates that in some circumstances, trying to track both targets causes one  to fail at both, and thus one is better off only trying to track one. The reason for this may be that a particular threshold amount of resource is needed to track a target, and so if neither target is allocated that much resource, tracking will fail for both. Evidence for this was provided by @chenResourceDemandsObject2013. In the terminology  of the @normanDatalimitedResourcelimitedProcesses1975 framework introduced in Chapter \@ref(whichAspects), the resource function that relates attentional resource proportion to accuracy falls below a straight line. This means that quantitative estimates of hemispheric independence will be overestimates, particularly in circumstances where the participants do not realize they may be better off focusing their efforts on tracking fewer targets than the number they have been told to track <!--FUTURE-->.

@carlsonQuadranticDeficitReveals2007 found evidence not only for hemifield independence but also quadrant-level independence, which they attributed to the partial separation of the retinotopic quadrant representations in areas V2 and V3. Using different stimuli, @shimSpatialSeparationTargets2008 and @holcombeObjectTrackingAbsence2014 did not, however, find evidence for quadrantic independence. Specifically, they did not find evidence for a deficit when two targets were positioned in the same quadrant compared to different quadrants but in the same hemifield. More work on this topic should be done.

## Some tracking resources are NOT hemifield-specific

One attentional process that is *not* hemifield-specific is feature attention, for example attention to color. When a participant is told to look for a red target, they are able to use feature attention to enhance all red objects, no matter where they are in the visual field [@whiteFeaturebasedAttentionInvoluntarily2011]. The decision to look for red originates with cognitive processes and remains hemifield-unified rather than hemifield-specific at the level of visual cortex [@saenzGlobalEffectsFeaturebased2002]. Indeed, people seem to be unable to confine the enhancement of red objects to one hemifield [@loHowWeSelect2014]. In real-world tracking where objects are at least somewhat heterogeneous and thus targets often have a different average color and other features than distractors, feature attention will facilitate tracking, and this facilitation is not hemifield-specific.

A previous Chapter (\@ref(Cequals1)) introduced the idea of C=1 cognitive processes that can support tracking of a single target but perhaps not multiple targets. Such processing likely is not hemisphere-specific, being aligned with "central executive" processes that integrate processing in both hemispheres. <!--We easily attend to objects in any part of the visual field, left or right.-->

@chenResourceDemandsObject2013 found evidence for both hemifield-specific tracking processes and processes not specific to a hemifield, operating in the same MOT task. Two targets were used, and on some trials they moved at different speeds. When a slow-moving target was paired (presented in the same trial) with a speedier target, accuracy was lower for the slow-moving target than if it was paired with a target that was slower. This suggests that participants allocate more tracking resources to the faster of two targets, presumably because slower targets do not require much resource to track well. This trade-off was most pronounced when the two targets were in the same hemifield, but seemed to occur to some degree even when the two targets were in different hemifields, implicating a cross-hemifield resource that plays a small role. This cross-hemifield resource may be a C=1 process. Furthermore, as discussed in the next section, perturbing one parietal lobe can affect performance in both hemifields, which suggests that each hemisphere can in some circumstances mediate tracking in either hemifield.

## The underlying mechanisms

The evidence reviewed above for hemifield independence suggests that hemisphere-specific processes determine how many targets one can track. This raises the question of what sort of processes those are, and how they interact with the cognitive processes that are more integrated across the hemispheres.

Steve Franconeri and colleagues have championed the idea that the hemisphere independence stems from spatial interference processes, by suggesting that these processes occur largely within a hemisphere [@franconeriNatureStatusVisual2013]. The idea is that when when an object is tracked, the neurons representing that target in retinotopic cortical areas activate inhibitory connections to nearby neurons in the cortical map, suppressing the responses to neighboring objects. To explain the findings of hemifield specificity, an additional detail of the account is the idea that the inhibitory neural connections do not extend from one hemisphere's retinotopic map to another [@franconeriFlexibleCognitiveResources2013]. This is plausible because in classic crowding tasks, spatial interference does show a discontinuity across the left- and right-visual field boundary lends some plausibility to this idea [@liuReductionCrowdingEffect2009]. However, @holcombeObjectTrackingAbsence2014 found evidence against spatial interference extending any further than the classic crowding range, which is only half the eccentricity of an object - for example, an object placed six degrees of visual angle from where the point the eyes are looking at would be interfered with only by other objects closer to it than three degrees of visual angle [@boumaInteractionEffectsParafoveal1970]. Because in most studies of hemifield independence, the stimuli are not close to the vertical midline, their would not be an artifactual finding of hemifield independence due to less crowding across the midline. The more viable theory of hemifield independence, then, is that of two neural resources that span each hemifield.

A number of studies have found that the activity of some parietal and frontal areas of cortex increase steadily with the number of targets in MOT [@culhamAttentionResponseFunctions2001; @howeUsingFMRIDistinguish2009; @jovicichBrainAreasSpecific2001; @alnaesPupilSizeSignals2014; @nummenmaaCorticalCircuitBinding2017], consistent with the importance of a pool of attentional resources. Unfortunately, these studies did not focus on the extent to which these activations are specific to target load within a hemifield, so we cannot be sure whether the brain activation measured reflects the hemifield-specific resource or a more global resource. The only imaging study I am aware of that investigated the issue is @shimNumberAttentionalFoci2010 who found an activation difference when the objects designated as targets were in opposite hemifields compared to when they were in the same hemifield. This was found for the superior parietal lobule and transverse parieto-occipital area, suggesting that they may be part of the hemifield-specific resource. The difference was not found for the anterior intraparietal sulcus, which could mean its activation reflects a global resource <!--FUTURE-->. 
<!--"In behavioral performance, participants tracked with higher accuracy when the 2 target pinwheels fell in opposite hemifields than in the same hemifield. A corresponding difference in brain activity was observed in the SPL (P < 0.05) and TrPO (P < 0.005), where activity was higher for track-2 unilateral than track-2 bilateral. The visual field arrangement did not affect activity in aIPS, however (P > 0.10), revealing a possible further dissociation in the PPC between the anterior and posterior" -->

@stormerWithinhemifieldCompetitionEarly2014 used EEG to investigate the hemifield-specific resource. They found that the SSVEP activation for targets was higher than that for distractors, especially when the two targets were positioned in different (left and right) hemifields. In contrast, an ERP component known as the P3 thought to reflect more cognitive identification and decision processes was similar in the two conditions. This is consistent with the theory that tracking depends on both hemisphere-specific attentive processing followed by some involvement of higher-order processes that are not hemisphere-specific.

@battelliRoleParietalLobe2009 found they could disrupt MOT performance in a hemifield by stimulating the contralateral intraparietal sulcus (IPS) using repetitive transcranial magnetic stimulation. Importantly, this only occurred when the moving targets were present in both hemifields. When the targets were all in the left or all in the right hemifield, TMS to the left or to the right IPS had no effect on tracking accuracy, and they replicated this in a second experiment. These findings bring to mind the competition between the two hemifields evident in the "extinction" symptom observed in parietal neglect patients. In extinction, responding to stimuli in the hemifield contralateral to parietal injury only shows significant impairment if there are also stimuli presented to the ipsilateral hemifield. This inspired Battelli to explain their findings with two propositions. The first is that the IPS in each hemisphere can mediate the tracking of targets in *either* visual hemifield. The second is that under normal conditions, inhibitory processes reduce the amount of ipsilateral processing by each IPS, causing tracking capacity to effectively be hemifield-specific.

A rather complicated relationship of the hemispheres is also suggested by evidence from patients. @battelliUnilateralRightParietal2001 found that in patients with damage to their right parietal lobe, MOT performance only in the left visual field was impaired relative to control participants. Evidently the right parietal lobe does not normally mediate tracking in the right visual field, so losing it did not hurt right visual field tracking performance. For another task, however, these right parietal patients had substantial impairments in *both* hemifields. Impairment on that task, an apparent motion task, is believed to be a result of a deficit for registering the relative timing of visual events. The involvement of the right parietal lobe, but not the left parietal lobe, in judging the temporal order of stimuli in both hemifields was further supported in an additional study with both patients and with TMS applied to intact brains [@agostaPivotalRoleRight2017].

In summary, while there is evidence that each parietal lobe is involved in field-wide processing for some tasks, parietal areas also likely mediate the hemifield independence evident in some circumstances. Using ERP, @drewSoftHandoffAttention2014 found evidence that when a target crosses the vertical midline, say from the left to the right hemifield, the left hemisphere becomes involved shortly before the target reaches the right hemifield, and the right hemisphere remains involved for a short time after the crossing. Because this was modulated by predictability of the motion, it did not appear to be wholly mediated by the well-known overlap of the two hemispheres' receptive fields at the midline. This phenomenon may reflect the normally-inhibited ipsilateral representation of the visual field by parietal cortices highlighted by @battelliRoleParietalLobe2009, although this remains uncertain as the location the ERP signals originated from was not clear.

Regarding the cooperation between the hemifields necessary to keep tracking a target that travels from one hemifield to another, both @strongHemifieldspecificControlSpatial2020 and @minamiHemifieldCrossingsMultiple2019 found evidence for a tracking performance cost when a target in MOT crossed the vertical midline. Evidently the handoff of control from one hemisphere to another is somewhat inefficient. @saikiRobustColorshapeBinding2019 also found some evidence in a memory paradigm that when two objects moved between hemifields, memory for their features was more disrupted than when they moved fron one to another quadrant within the same hemifield. @strongHemifieldspecificControlSpatial2020 found no cost when targets moved between quadrants while remaining within a hemifield, an important finding given that other work raised the prospect of quadrant-specific resources [@carlsonQuadranticDeficitReveals2007]. 

In summary, areas of the parietal cortex may subserve both the hemifield-specific tracking resource that dominates most MOT tasks and the possible resource that is not specific to a hemifield. More work must be done, however, to determine the role of frontal lobe regions. Such regions could potentially play a role in the hemifield-specific resource, the hemifield-independent resource, or both.

## What else are hemifield-specific resources used for?

Multiple object tracking appears to be much like the spatial selection studied extensively in the visual attention literature, just applied to moving objects. Thus MOT suggests hemifield specificity of spatial attentional selection, and one should expect to find strong hemifield-specificity of other visual cognition tasks in addition to MOT. 

Prior to the discovery that MOT is highly hemifield-specific, many researchers had already, for various tasks, compared performance when the two stimuli are presented in the same hemifield to performance when they are presented in different hemifields. For example, @dimondUseTwoCerebral1971 found that the reporting of two briefly-presented digits is more accurate when the digits are presented in different hemifields than in the same hemifield. However, this study and many like it did not include a single-stimulus condition, so for the higher performance in the split condition, we don't know how close that is to the one-target level of performance and thus the degree of different-hemifield advantage cannot be quantified. Moreover, many studies use response time as a measure, which can be difficult to interpret quantitatively [@awhEvidenceSplitAttentional2000; @serenoDiscriminationHemifieldsNew1991; @dimondUseTwoCerebral1971]. 

<!-- A' is no good https://sites.google.com/a/mtu.edu/whynotaprime/ -->
@delvenneCapacityVisualShortterm2005 used both dual-target and single-target conditions in a visual working memory task. For spatial working memory, he estimated 40% hemifield independence, although unfortunately he used the discredited A' measure of performance [@zhangNoteROCAnalysis2005] and did not space the stimuli widely enough to reduce the possibility of spatial interference. Nevertheless, the advantage appears to be large and did not occur for color working memory [@delvenneVisualShorttermMemory2012]. More generally, only tasks with spatial demands seem to show much of a different-hemifield advantage [@holtBilateralAdvantageMaintaining2015; @umemotoBilateralAdvantageStorage2010]

@alvarezAnatomicalConstraintsAttention2012a studied visual search, with the stimuli to search arrayed bilaterally or unilaterally. They found only a small advantage of the vertical meridian split in a standard search task. However, in a subset search task where participants knew the target would be located in one of several locations designated by a pre-trial cue, they found a large bilateral advantage. When the relevant locations were visually salient (due to a color difference), this advantage largely disappeared. These results, and those in the rest of the literature, suggest that hemifield advantages are strongest when spatial selection is critical.

@strongHemifieldspecificControlSpatial2020 investigated working memory for stimuli that moved either within a hemifield or between hemifields. For between-hemifield movement, they found a substantial decrease in accuracy for the spatial task of remembering which positions of a 2x2 grid contained dots at the beginning of the trial, before the (empty) grid moved - 79% correct for between-hemifield movement, and 85% correct for within-hemifield movement. No such cost was found for color or identity memory tasks. This between-hemifield cost for spatial memory was similar to the cost they found for MOT itself.

This association between spatial tasks and a different-hemifield advantage may reflect a large-scale difference in how the brain processes spatial versus identity information. Famously, the dorsal stream to the parietal cortices are more concerned with spatial information than is the ventral stream, which is more involved in object recognition [@goodaleSeparateVisualPathways1992]. Neural responses in the dorsal pathway to parietal cortex are largely contralateral [@serenoMappingContralateralSpace2001], although as we have seen, this may depend on having stimuli in both hemifields. Largely contralateral responding is also found for other brain areas thought to contribute to a "saliency map" [@fecteauSalienceRelevanceFiring2006], such as the frontal eye fields [@haglerjrSpatialMapsFrontal2006], the superior colliculus [@schneiderVisualResponsesHuman2005], and the pulvinar [@cottonContralateralVisualHemifield2007a]. In contrast, identity-related processing seems to involve more bilateral neural responses and connectivity between hemispheres [@cohenUsingNeuronalPopulations2011; @hemondPreferenceContralateralStimuli2007].

Multiple identity tracking, which is discussed further in section \@ref(identity), combines the location-updating demand of multiple object tracking with an additional requirement to maintain knowledge of what features belong to each of the objects. Across four experiments, @hudsonHemifieldEffectsMultiple2012 consistently found partial hemifield independence for this task, ranging from 26 to 37% (my calculations are [here](https://github.com/alexholcombe/tracking-review/blob/main/twoBrainsOrOne.Rmd)) with a paradigm that yielded `r round(100*HudHoweLittleMOTindependence)`% independence for MOT. This is consistent with the suggestion of the findings listed above that spatial selection and/or location updating processes are much more hemisphere-specific than processes that require maintenance of non-spatial features.

<!--
@delvenneBilateralFieldAdvantage2011 showed a bilateral advantage for subitizing.  Subitizng and enumeration. Bilateral advantage shown for enumeration. None for subitizing, although that's hard to interpret because ceiling effect 

@delvenneVisualShorttermMemory2012 review article: "Those data reveal that a BFA can be found in spatial short-term memory, namely when the task requires holding spatial information. However, no BFA has been found in object short-term memory (i.e., when the task is to hold detailed visual properties of the objects), unless attentional selective processes are dominantly engaged in the task, s"

@umemotoBilateralAdvantageStorage2010 found bilateral advantage even with sequential so must be storage, althoguh used orientation, so could be spatial still
used mixture model, but  might have assessed capacity quantitatively Mixture model suggests only a modest increase, from perhaps capacity of 2 . 0.5*4, Figure 11 . to capacity of 2.3 .575*4, Figure11

Kingstone, 2004; Corballis & Gratton, 2003; Hughes & Zimba, 1987

RT study, always used 2 targets.  Davis, R., & Schmit, V. (1971). Timing the transfer of information between hemispheres in man. Acta Psychologica, 35(5), 335–346. https://doi.org/10.1016/0001-6918(71)90008-4
-->

## Hemispheric differences

MOT and spatial selection appear to be limited by independent processing in the two hemispheres. Are these two hemispheres doing exactly the same thing, or do they have different strengths?

Functional differences between the left and right cerebral hemispheres can be attenuated at the behavioral level by the cross-hemisphere integration that occurs for many tasks prior to the stage that determines behavioral performance. The fact that tracking performance reflects more hemispheric independence, then, provides a higher potential to show hemifield differences than for most tasks. As it turns out, however, while differences are observed, they do not seem to be large.

In each of four experiments conducted in 2013, we found either a trend for or statistically significant advantage for targets in the right hemifield (Figure A2) [@holcombeObjectTrackingAbsence2014]. This was also found by @strongHemifieldspecificControlSpatial2020. <!--Interestingly, there was some non-significant evidence that this was greater in their one-target condition than in their two-target condition.    Figure 4 of @battelliUnilateralRightParietal2001 like Holcombe et al. (2014) finds (non-significant) better tracking performance in RVF than the left only when tracking one rather than more than one--> The right hemifield advantage could be explained by the idea that stimuli presented to the right hemifield are processed by both hemispheres to a greater degree than are stimuli presented to the left hemifield — this is thought to be why left neglect is more common than right neglect. Specifically, it is thought that the right hemisphere mediates attention to *both* hemifields [@mesulamSpatialAttentionNeglect1999], such that the right hemifield is doubly represented. However, while @strongHemifieldspecificControlSpatial2020 confirmed a right hemifield advantage in their MOT experiments, they found a *left* visual field advantage for spatial working memory experiments, even though spatial working memory is also thought to be mediated by parietal cortex. Most strikingly, @matthewsLeftVisualField2015 found a large advantage for temporal order judgments and simultaneity judgments for stimuli presented to the left hemifield. Neuropsychological evidence suggests that the left hemifield advantage for timing tasks reflects a specialization for timing in the right parietal cortex [@battelliBilateralDeficitsTransient2003], even though human specialization for language, which is thought to be in part an enhancement of timing and sequencing abilities, has occurred more in the left hemisphere.

The situation becomes more complex once one considers that subtle interactions between the two hemispheres seem to affect attention in each hemifield, as highlighted in the "the underlying mechanisms" section above. One illustration of this is a recent finding by @edwardsBehavioralGainFollowing2021. They had participants perform MOT in either hemifield for a 30 min session, and afterwards found that performance in the *untrained* hemifield improved significantly. The reason for this is not clear, but could reflect "fatigue" by the hemisphere contralateral to the trained hemifield, causing it to reduce its inhibition of the other hemisphere. Alternatively, the mechanism could be potentiation (an increase in gain) of the untrained hemisphere as a result of the deprivation, which may be the reason why depriving an eye results in increased cortical activity when that eye is stimulated later [@lunghiBriefPeriodsMonocular2011].





